A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-09-26 paper/project.tex
Edits to project figures and text. Tests skipped!

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-298}
\end{center}
\caption{Liquid-vapor interface profile at room temperature.  }
\label{fig:liquid-vapor-profile}
\end{figure}

@@

For reference, we plot the predicted vapor pressure (see
Figure~\ref{fig:equation-of-state}) and coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) against experimental values.  These
match those calculated by Clark \emph{et al}. To find the densities, we use the
``common tangent'' rule, as illustrated in Figure~\ref{fig:homogeneous}.
\needsworknow{Expand the previous paragraph}

@@

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the \emph{exact}
density-functional theory \emph{would} include them), with the result
that the density profile will be smoothed out by CW effects. The liquid-vapor
interface is plotted in Figure~\ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profiles illustrating the transition from vapor 
to liquid water between the rods. The radius is 0.5~nm, the top figure is at a separation of 0.2~nm and the
bottom is 0.4~nm. }
\label{fig:density-rods}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods ranging in radius from
0.3~nm to 0.9~nm. All were arbitrarily set to go to zero at large separations. The
discontinuous transition corresponds to the phase change from
vapor to liquid between the rods. }
\label{fig:rods-energy-vs-distance}
\end{figure}

@@

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of various radii at separation $d$ (see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to about 1~nm. At small
separations ($d<2$~-~4~nm) there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. Since we are looking at only hard-walled
non-interacting rods, the critical separation for this transition will be
different than any system where there is an appreciable force between rods.

The free energy per length is calculated for several different radii and is plotted
in Figure~\ref{fig:rods-energy-vs-distance}. Each data set is adjusted such that the
free energy per length goes to zero at large separations. Once a certain separation is
reached, the rods will simply act like the single rod in water and the energy will
reach an asymptote proportional to the surface tension (as in Figure~\ref{fig:energy-vs-diameter}).

For one set of rods (radii of 0.5~nm) the free energy is plotted with the internal energy
and the entropic energy contribution, where all are in units of energy per length 
(see Figure~\ref{fig:energy-rods}). \needsworknow{More discussion of Figure~\ref{fig:energy-rods}}

Walther
\emph{et al.}\cite{walther2004hydrodynamic} study the interactions between two
carbon nanotubes. They use molecular dynamics with Lennard-Jones potentials for
nanotubes of diameter 1.25~nm and separations ranging from about 0.3~nm to 1.5~nm.
They also find a drying transition occurs, and find it to be at separations less
than 0.9~-~1.0~nm\cite{walther2004hydrodynamic}. This is several times larger than
what we find for hydrophobic rods, as expected due to the carbon-carbon
interations between two nanotubes.

The density oscillations are very similar to
those for the single hydrophobic rod in water, as seen in Figure~\ref{fig:density-single-rod}.

@@

[conclusion text]

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-298}
\end{center}
\caption{The liquid-vapor density profile at room temperature shows
some small oscillations close to the interface.  }
\label{fig:liquid-vapor-profile}
\end{figure}

@@

For reference, we plot the coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) and the predicted vapor pressure (see
Figure~\ref{fig:equation-of-state}) against experimental values.  These
match those calculated by Clark \emph{et al}. The functional provides a good description of
the vapor pressure, which has an error of less than 
1\% of the experimental data. The coexistence curve is calculated by 
\fixme{explanation of coexistence curve calculation?}. Near the critical temperature,
the agreement is off, and the theoretical prediction  of the temperature is quite a bit higher than
experiment. To find the densities, we use the
``common tangent'' rule, as illustrated in Figure~\ref{fig:homogeneous}.
\needsworknow{Expand the previous sentence - common tangent rule}

@@

@@

The fluid density at a liquid-solid interface commonly displays
oscillatory behavior, but the presence of density oscillations at
liquid-vapor interfaces is more debated\cite{penfold2001structure}.
In fact, the \emph{meaning} of the density profile at a liquid-vapor
interface is clouded by the existence of long-wavelength perturbations
known as \emph{capillary waves}.  These waves are not accounted for in
density-functional theory (although the exact
density-functional theory would include them), with the result
that the density profile will be smoothed out by CW effects. The liquid-vapor
interface is plotted in Figure~\ref{fig:liquid-vapor-profile} and shows
some small oscillations.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-rods-in-water}
\end{center}
\caption{ Density profiles illustrating the transition from vapor 
to liquid water between the rods. The radius is 0.5~nm, the top figure is at a separation of 0.2~nm and the
bottom is 0.4~nm. Figures~\ref{fig:rods-energy-vs-distance}~and~\ref{fig:energy-rods} show
the energy for these and other separations.}
\label{fig:density-rods}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods ranging in radius from
0.3~nm to 0.9~nm. All were arbitrarily set to go to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-rods}. }
\label{fig:rods-energy-vs-distance}
\end{figure}

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of various radii at separation $d$ (see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to about 1~nm. At small
separations ($d<2$~-~4~nm) there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. This critical separation for the transition to liquid depends
somewhat on the radii of the rods, and is about 0.25~nm for the rods in Figure~\ref{fig:density-rods}.
Since we are looking at only hard-walled
non-interacting rods, the critical separation for this transition will be
different than any system where there is an appreciable force between rods. 
The density oscillations  around the rods are very similar to
those for the single hydrophobic rod in water, as seen in Figure~\ref{fig:density-single-rod}.
The oscillations around the two rods at small separation looks as if the two rods were one
solid object shaped \fixme{racetrack shaped?}.

For one set of rods (radii of 0.5~nm) the free energy is plotted with the internal energy
and the entropic energy contribution, where all are in units of energy per length 
(see Figure~\ref{fig:energy-rods}). The free energy slightly increases as the separation
goes from 0 to the critical separation (0.25~nm) and then is approximately constant at
larger separations. Since there is no repulsive interaction between the rods, there is no
large energy increase when the rods touch. The entropy has a minimum just before the rods
transition to let liquid between them. As the rods separate starting from zero, the area 
between them containing vapor increases and there is less overall volume for the 
surrounding liquid water to occupy and thus the entropy decreases. 
\needsworknow{More discussion of Figure~\ref{fig:energy-rods}, internal energy, how they add
together, etc.}

The free energy per length is compared for rods of several different radii
in Figure~\ref{fig:rods-energy-vs-distance}. For comparison, we arbitrarily adjust each data set such that the
free energy per length goes to zero at large separations. At large separations, the rods will 
simply act like two separate single rods in water and the total energy will
reach an asymptote proportional to the surface tension (similar to that shown in 
Figure~\ref{fig:energy-vs-diameter}).
The change in free energy per length at the point of transition increases as the radii of
the rods increases. The smallest radius we show is 0.3~nm, which shows only a slight increase in
free energy at the critical separation, while the largest radius has a more significant jump in free energy. 
We see that all the rods studied have the same behavior, with a positive linear relation between free energy
per length and the separation during the initial state where the rods have vapor between them. 
The slope of this line is equal to the force per length on the rods, which we calculate to be 144~mN/m, 
or twice the surface tension
found by the functional (see Figure~\ref{fig:surface-tension}).

Walther \emph{et al.}\cite{walther2004hydrodynamic} study the interactions between two
carbon nanotubes, which are very geometrically similar to our hydrophobic rods. 
They use molecular dynamics with Lennard-Jones potentials for
nanotubes of diameter 1.25~nm and separations ranging from about 0.3~nm to 1.5~nm.
They also find a drying transition occurs, and find it to be at separations near
0.9~-~1.0~nm\cite{walther2004hydrodynamic}. This is several times larger than
what we find for hydrophobic rods, as expected due to the carbon-carbon
interations between two nanotubes.

@@

\needsworknow{Conclusions about slit, single rod, two hydrophobic rods (mostly). future uses}
